Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-07 manuscript/manuscript.tex
more Gs changes

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under strong background selection is invariant
with respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure that seems reasonably invariant to the choice
of $s$ (Supplementary Material Figure \ref{suppfig:supp-bgs-g}). We believe
that these observations can be reconciled by $Cov(\Delta p_t,~ \Delta
p_{t^{\prime}})$ being larger for larger $s$ when $|t^{\prime}-t|$ is small,
but also decaying more rapidly with $|t^{\prime}-t|$, such that the overall
contribution of selection to allele frequency variance is invariant to $s$ (for
strong background selection). However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{figures/fig-bgs-G-without-fixations.pdf}

  \caption{The trajectories of $G(t)$ through time under background selection,
    under different recombination rates ($r_{BP}$, rows), selection
    coefficients ($s$), and deleterious mutation rates ($U$). The first column
    sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
  constant, and $U$ varies.}

  \label{suppfig:supp-bgs-g}
\end{figure}

@@

While investigating} whether different selection coefficients converged to
the same levels of $G(t)$, we extended the number of generations we calculated
our statistics on (e.g. in earlier exponential and Gaussian stabilizing
selection fitness function, these were calculated to fifty generations; here
we've calculated them to one hundred generations). This lead us to discover a
subtle bias temporal covariance: whether sites that fix or are lost through
time are included or excluded in the temporal covariance calculation can lead
to biases in temporal covariance. We discuss this extensively in the next
section, Supplementary Materials Section \ref{supp:fixation}, but here we note
that all the background selection results exclude fixed and lost sites; these
are marked as missing and excluded from the temporal covariance and $G(t)$
calculations.

@@

\subsubsection{The Effect of Fixations in the empirical datasets}
\label{supp:fixation}

In analyzing both the empirical datasets included in our study and our
simulation results, we noticed the temporal covariances and $G(t)$ statistics
can differ depending on how allele frequencies of zero or one are handled.
Generally, temporal covariances should be calculated on polymorphic sites; once
a site has reached fixation or loss, its allele frequency change $\Delta p_t =
0$ and including these sites in the temporal covariance calculation can lead to
biases, which we discuss below. However, with \emph{sample} allele frequencies,
rather than population frequencies, a site with observed frequency zero or one
may still be segregating, but by chance not sampled at a timepoint. Here, we
discuss the effect of including sites with frequency zero or one, and show our
empirical results are not qualitatively different when analyzed excluding fixed
sites.

  \caption{$G(t)$ trajectories from background selection simulations both
  including fixed sites (solid lines) and not including fixed sites (dashed
lines). The dashed lines are identical to those in Supplementary Materials
Figure \ref{suppfig:supp-bgs-g}. }

We noticed in some simulations, when temporal covariances are calculated over
longer time periods (i.e. $>50$ generations), that neutral simulations appear
to slowly trend towards negative $G(t)$ (Supplementary Materials Figure
\ref{suppfig:supp-bgs-g-fix}). In investigating this bias, we found that the
effect went away when we marked fixed or lost sites as missing, and calculated
the temporal covariance; furthermore, this affected the $G(t)$ trajectories
under selection too (Supplementary Material Figure
\ref{suppfig:supp-bgs-g-fix}).

Looking a neutral simulations results, we found that spuriously significant
$G(t)$ trajectories (e.g. different from zero under the case of neutrality) can
be created by conditioning on sojourn time. Our intuition, which will need to
be confirmed by future work and theory, is that by conditioning on a quick
sojourn time to fixation (or loss) from an initial intermediate frequency,
leads to trajectories that, purely by chance, have many upward (or downward)
allele frequency changes, which leads to a positive sum of pairwise temporal
covariances (the numerator of $G(t)$). Similarly, conditioning on long sojourn
times implies that frequency changes alternate back and forth quite a bit, such
the sum of pairwise temporal covariances could be slightly negative. However,
the overall affect could also be affected by the average allele frequency
change across loci and other factors.

To further complicate the matter, our simulation allele frequencies are
population, rather than sample frequencies, when an allele is fixed or lost is
known with certainty. With empirical data calculated on sample allele
frequencies, low frequency minor alleles many not be sampled at some
timepoints, and excluding these observations (instead of treating it as a
trajectory that has a 0 frequency timepoint) biases estimates.  We observed
this bias created by ignoring fixed or loss sites by comparing our $N_e$
estimates for the \textcite{Barghi2019-qy} study (estimated from the total
variance in allele frequency change) with the estimates from the original
paper. We found that ignoring fixed or lost sites (by treating them as missing
data in calculating the covariances) led to many low-frequency alleles not
contributing to the variance, leading this statistic to be calculated on more
intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
Additionally, we tried only dropping fixed or loss sites from the temporal
covariance calculations that were at the end or the beginning of a trajectory
(e.g., as if the site was created by a new mutation or fixed); while this
ameliorated some of this bias, it still did not lead to $N_e$ estimates
congruent with the original studies. Overall, we found by trying all these
approaches that not removing fixed or lost sites was the best way to deal with
sample allele frequencies that could be missing from some timepoints.

>>>>

\gc{COMMENT ON LARGER U than usual}

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under strong background selection is invariant
with respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure that seems reasonably invariant to the choice
of $s$ (Supplementary Material Figure \ref{suppfig:supp-bgs-g-fix}). We believe
that these observations can be reconciled by $Cov(\Delta p_t,~ \Delta
p_{t^{\prime}})$ being larger for larger $s$ when $|t^{\prime}-t|$ is small,
but also decaying more rapidly with $|t^{\prime}-t|$, such that the overall
contribution of selection to allele frequency variance is invariant to $s$ (for
strong background selection). However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

  \caption{The trajectories of $G(t)$ through time under background selection,
    under different recombination rates ($r_{BP}$, rows), selection
    coefficients ($s$), and deleterious mutation rates ($U$). The first column
    sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
  constant, and $U$ varies. Calculated using both
  including fixed sites (solid lines) and not including fixed sites (dashed
lines). \gc{MOVED THIS FIG UP} }

@@

%\begin{figure}[!ht]
 % \centering
  %\includegraphics[width=\textwidth]{figures/fig-bgs-G-without-fixations.pdf}

  %\caption{The trajectories of $G(t)$ through time under background selection,
   % under different recombination rates ($r_{BP}$, rows), selection
  %  coefficients ($s$), and deleterious mutation rates ($U$). The first column
  %  sets $U = 1.0$, and $s$ varies, while in the second column $s=0.05$ is held
 %% constant, and $U$ varies.}

 % \label{suppfig:supp-bgs-g}
%\end{figure}

@@

\gc{We found that the the results for background selection varied
  according to whether sites that fix or are lost through
time are included or excluded in the temporal covariance calculation can lead
to differences in the estimated $G(t)$temporal covariance. We discuss
this extensively in a subsequent section, Supplementary Materials Section \ref{supp:fixation}.}

%While investigating} whether different selection coefficients converged to
%the same levels of $G(t)$, we extended the number of generations we calculated
%our statistics on (e.g. in earlier exponential and Gaussian stabilizing
%selection fitness function, these were calculated to fifty generations; here
%we've calculated them to one hundred generations). This lead us to discover a
%subtle bias temporal covariance: whether sites that fix or are lost through
%ime are included or excluded in the temporal covariance calculation can lead
%to biases in temporal covariance. We discuss this extensively in the next
%section, Supplementary Materials Section \ref{supp:fixation}, but here we note
%that all the background selection results exclude fixed and lost sites; these
%are marked as missing and excluded from the temporal covariance and $G(t)$
%calculations.

@@

\subsubsection{The Effect of Fixations in the empirical datasets}
\label{supp:fixation}
\gc{In our simulation results, we noticed the temporal covariances and $G(t)$ statistics
can differ depending on how allele frequencies of zero or one are handled.

%In analyzing both the empirical datasets included in our study and our
%simulation results, we noticed the temporal covariances and $G(t)$ statistics
%can differ depending on how allele frequencies of zero or one are handled.
Generally, temporal covariances should be calculated on polymorphic sites; once
a site has reached fixation or loss, its allele frequency change $\Delta p_t =
0$ and including these sites in the temporal covariance calculation can lead to
biases. However, with \emph{sample} allele frequencies,
rather than population frequencies, a site with observed frequency zero or one
may still be segregating, but by chance not sampled at a timepoint. Here, we
discuss the effect of including sites with frequency zero or one, and show our
empirical results are not qualitatively different when analyzed excluding fixed
sites.

With empirical data calculated on sample allele
 frequencies, low frequency minor alleles many not be sampled at some
 timepoints, and excluding these observations (instead of treating it as a
 trajectory that has a 0 frequency timepoint) biases estimates of
 quantities such as $N_e$ as it biased the analysis towards
 intermediate frequency alleles.
 Additionally, we tried only dropping fixed or loss sites from the temporal
 covariance calculations that were at the end or the beginning of a trajectory
 (e.g., as if the site was created by a new mutation or fixed); while this
 ameliorated some of this bias it did not remove all of it. 
 Overall, we found by trying all these approaches that not removing fixed or lost sites was the best way to deal with
 sample allele frequencies that could be missing from some timepoints. }

% To further complicate the matter, our simulation allele frequencies are
% population, rather than sample frequencies, when an allele is fixed or lost is
% known with certainty. With empirical data calculated on sample allele
% frequencies, low frequency minor alleles many not be sampled at some
% timepoints, and excluding these observations (instead of treating it as a
% trajectory that has a 0 frequency timepoint) biases estimates.  We observed
% this bias created by ignoring fixed or loss sites by comparing our $N_e$
% estimates for the \textcite{Barghi2019-qy} study (estimated from the total
% variance in allele frequency change) with the estimates from the original
% paper. We found that ignoring fixed or lost sites (by treating them as missing
% data in calculating the covariances) led to many low-frequency alleles not
% contributing to the variance, leading this statistic to be calculated on more
% intermediate frequency alleles and thus biasing the $N_e$ estimate downwards.
% Additionally, we tried only dropping fixed or loss sites from the temporal
% covariance calculations that were at the end or the beginning of a trajectory
% (e.g., as if the site was created by a new mutation or fixed); while this
% ameliorated some of this bias, it still did not lead to $N_e$
% estimates  GUESS I"M UNSURE WHY THIS IS A CRITERIA FOR BEING CORRECT
% congruent with the original studies. Overall, we found by trying all these
% approaches that not removing fixed or lost sites was the best way to deal with
% sample allele frequencies that could be missing from some timepoints.
